---
title: "Master Thesis Defence"
author: "Malte Berneaud-Kötz"
date: "May 19, 2016"
output:
  beamer_presentation:
    keep_tex: yes
    colortheme: beaver
    fonttheme: structurebold
    slide_level: 2
    theme: Malmoe
subtitle: The Eurosceptic Misfit - Popular Euroscepticism and Electoral Support for
  Eurosceptic Parties
header-includes:
- \usepackage{graphicx}
- \usepackage{tikzpagenodes}
- \usetikzlibrary{calc}
- \usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE, cache = TRUE, message = FALSE, fig.path = "graphs/",
                      warning = FALSE, fig.align = "center", dev = "pdf")
```

```{r analysis}
source("../../master.R")
library(gridExtra)
```


## Overview

1. Motivation
    + The phenomenon of the "Eurosceptic misfit"
    + A gap in the research field
2. Methodology
    + The data set
    + The statistical models
3. Findings
    + Eurosceptic misfit across time
    + Eurosceptic misfit across regions
    + Regression models
4. Conclusion

# 1. Motivation
## The phenomenon of the "Eurosceptic misfit"

+ Wide gap between popular levels of Euroscepticism and the aggregate vote share received by Eurosceptic parties
+ First mentioned by Taggart (1998)
    + He suspected the translation of Eurosceptic attitudes into Eurosceptic vote was driven by national contextual factors  
  
**Understanding national contextual drivers of the "Eurosceptic misfit" is a puzzle-piece in understanding the political impact of Eurosceptic parties**

## A gap in the research field

+ Literature on Euroscepticism mainly revolves around:  
    + Categorisation of Euroscepticism
    + Derterminants of Euroscepticism
+ The Eurosceptic misfit has received no thorough scholarly attention outside a few mentions (e.g. Taggart & Sczerbiak, 2002; Verney (2011))

**Possibility to make a contribution to an underdeveloped research field**

# 2. Methodology

## 2.1. The data set
I created my own study data set from

  + European Parliament election data from 1979 to 2009
  + Eurobarometer data adjacent to each election round

+ This yielded a data set which had
    + 110 observations  
    + 27 panel units  
    + 7 time periods  

## 2.2. The statistical models

+ **Dependent variable**: share of voters holding Eurosceptic attitudes - sum of the vote share Eurosceptic parties  

+ **Independent variables**

    + Two measures of Eurosceptic attitudes
    + Party system polarisation
    + Effective number of parties
    + Membership duration in the EU
    + Location in Central and Eastern Europe

+ Models
    + Fixed effects, random effects (and pooled OLS)


# 3. Findings
## 3.1. Eurosceptic misfit across time
```{r graph, include=TRUE, echo=FALSE, warning=FALSE, fig.height=3, fig.width = 4}
ggplot(year_averages, aes(x = year)) +
              geom_line(aes(y = EUS.avg, colour = "EUS.avg")) +
              geom_line(aes(y = vote.avg, colour = "vote.avg")) +
              geom_line(aes(y = diff.avg, colour = "diff.avg")) +
scale_colour_discrete(name = "Euroscepticism variables", 
                                 labels = c("Eurosceptic misfit", 
                                            "Popular Euroscepticism",
                                            "Eurosceptic vote share")) +
  ylab("Variable value in %") + xlab("Year of EP election") + 
  scale_x_continuous(breaks = c(1979, 1984, 1989, 1994, 1999, 2004, 2009)) +
  geom_vline(xintercept = c(1981, 1995, 2004, 2007), alpha = 0.3) +
  theme(legend.position="top", legend.title = element_blank(), 
        legend.text = element_text(size = 6),
        text = element_text(size = 6), axis.title.x = element_text(size = 7), 
        axis.title.y = element_text(size = 7))
```

## 3.2. Eurosceptic misfit across regions 
```{r creating_plots}
NE.p <- ggplot(NE, aes(y = diff.EUS, x = cname)) + geom_bar(stat = "identity") +
  ylab("Average Eurosceptic Misfit") + xlab("") +
  expand_limits(y = c(-3, 28))+
  theme(text = element_text(size = 5)) + ggtitle("Northern Europe")

WE.p <- ggplot(WE, aes(y = diff.EUS, x = cname)) + geom_bar(stat = "identity") +
  ylab("Average Eurosceptic Misfit") + xlab("")+
  expand_limits(y = c(-3, 28))+
  theme(text = element_text(size = 5)) + ggtitle("Western Europe")

SE.p <- ggplot(SE, aes(y = diff.EUS, x = cname)) + geom_bar(stat = "identity") +
  ylab("Average Eurosceptic Misfit") + xlab("")+ 
  expand_limits(y = c(-3, 28))+
  theme(text = element_text(size = 5)) + ggtitle("Southern Europe")

CEE.p <- ggplot(CEE, aes(y = diff.EUS, x = cname)) + geom_bar(stat = "identity") +
  ylab("Average Eurosceptic Misfit") + xlab("")+
  expand_limits(y = c(-3, 28))+
  theme(text = element_text(size = 5)) + ggtitle("Central/Eastern Europe")
```

```{r displaying_plot, include=TRUE, echo=FALSE, fig.height=3, fig.width = 4}
grid.arrange(NE.p, WE.p, SE.p, CEE.p, nrow = 2)
```

## 3.3. Regression models
```{r regression results, include=TRUE, echo=FALSE, warning=FALSE, results='asis'}
stargazer(robust.fe, robust.re, title = "Regression results",  column.labels = c("FE", "RE"),
          dep.var.labels = "Eurosceptic Misfit", covariate.labels = 
            c("General Euroscepticism", "Instrumental Euroscepticism",
              "Polarisation Index", "Effective Number of Parties", 
              "Membership Duration", "Central/Eastern European"), digits = 2,
          header = FALSE, font.size = "tiny")
```

## 4. Conclusion

+ The Eurosceptic misfit has stayed relatively constant over time, except for a spike in 1994
+ Eurosceptic misfit somewhat smaller in CEE countries
+ Eurosceptic vote share generally fails to "catch up" with increases in popular Eurosceptic attitudes
+ Higher degrees of party polarisation shrink the Eurosceptic misfit

## Thank you for listening!

Please check out my [Github repository](https://github.com/mberneaud/EuroscepticMisfitMasterThesis) for all source codes for the analysis and presentation documents:  

_github.com/mberneaud/EuroscepticMisfitMasterThesis_

## Extra: Model considerations 

+ Combination of pooled OLS, fixed effects and random effects models used in the thesis
    + Fixed effects to cleanly isolate causal effects (read: removing unobserved heterogeneity) to measure country-specific variation
    + Random effects to allow for inclusion of time-invariant variables (like region) and higher efficiency
    + Pooled OLS for comparison